---
title: "10_site_study_data_filtering"
author: "Melissa Carew"
date: "21/08/2024"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


This notebook is designed to filling missing taxonomic assignments from the vsearch step (in 03_classification) in the metabarcoding workflow and make corrections to taxonomic assignments that are incorrectly assigned by vsearch due to errors in the DNA barcode reference libraries (SECTION 1).It also performs a final flitering to detect and remove species detections that have most likely arisen from 'mistagging' (SECTION 2). It has been modified to produce a final dataset for the 10 site study.


```{r}
# Load the required libraries
library(dplyr)
library(tidyr)
library(stringr)
library(readxl)
library(here)
```

# reading in data
The MiSeq13 runs contains the bulk of the data for the 10 site study. However, a few samples require re-running therefore the data need to be retrieved and combined into a final dataset prior to filtering.

Samples requiring re-running
40BNY29042 <- re-run in MiSeq15 (3 replicates 40BNY29042rep3, 40BNY29042rep4 and 40BNY29042rep5 (only need 2 reps))

10FEH41841 <- re-run in MiSeq15 (3 replicates labelled 10FEH41841rep3, 10FEH41841rep4 and 10FEH41841rep5 (only need 2 reps))

30DNG75620 <- re-run in MiSeq16 (3 replicates labelled Elut30DNG79574rep1, Elut1030DNG79574rep2 and Elut30DNG75620rep3 (only need 2 reps))

```{r}
file_path <- here::here("results/final_data_summaries/")

# Set the directory path for files to be stored on uomshare (NOTE: 'test' will need to be change to your miseq folder name)

#read in final vsearch data summaries
miseq13_vsearch_sum <- read.csv(paste0(file_path, "miseq13_vsearch_data_summary_23_02_2026.csv"))
miseq15_vsearch_sum <-read.csv(paste0(file_path,"miseq15_vsearch_data_summary_26_11_2024.csv"))
# fix a small error in dataframe
miseq15_vsearch_sum <- miseq15_vsearch_sum %>%
  mutate(Consensus = ifelse(asv_code == '956bcb88ea0e4762eefac3c2a93bbb2f', 0.857, Consensus))
miseq16_vsearch_sum <- read.csv(paste0(file_path, "miseq16_vsearch_data_summary_26_11_2024.csv"))
```

```{r}
# remove columns to be replaced re-run data from MiSeq13                               
miseq13_vsearch_sum_mod <- miseq13_vsearch_sum[, !(names(miseq13_vsearch_sum) %in% c("X10FEH41843rep1", "X10FEH41843rep2", "X40BNY29042rep1", "X40BNY29042rep2", "X30DNG79574rep1", "X30DNG79574rep2"))]

# subset rerun data from Miseq15 and Miseq16 to merge with miseq13_vsearch_sum_mod
miseq15_vsearch_reruns <- miseq15_vsearch_sum %>% select(asv_code, kingdom, phylum, class, order, family, genus, species, Consensus, max_p_identity, threshold, X40BNY29042rep3, X40BNY29042rep4, X10FEH41841rep3, X10FEH41841rep4, DrosXrep1, DrosXrep2, DrosXrep2, MAR2Trep1, MAR2Trep2, MAR2Trep3, asv_seq, amplicon)
# Rename sample columns to site names
miseq15_vsearch_reruns <- miseq15_vsearch_reruns %>%
  dplyr::rename(X10FEH41843rep3 = X10FEH41841rep3, X10FEH41843rep4 = X10FEH41841rep4)

miseq16_vsearch_reruns <- miseq16_vsearch_sum %>% select(asv_code, kingdom, phylum, class, order, family, genus, species, Consensus, max_p_identity, threshold, Elut30DNG79574rep1, Elut30DNG79574rep3, DrosXrep1, DrosXrep2, DrosXrep3, MAR2Trep1, MAR2Trep2, MAR2Trep3, asv_seq, amplicon)
# rename sample name to site name
miseq16_vsearch_reruns <- miseq16_vsearch_reruns %>% dplyr::rename(`X30DNG79574rep1` = Elut30DNG79574rep1, `X30DNG79574rep2` = Elut30DNG79574rep3, DrosXrep4 = DrosXrep1, DrosXrep5 = DrosXrep2, DrosXrep6 = DrosXrep3, MAR2Trep4 = MAR2Trep1, MAR2Trep5 = MAR2Trep2, MAR2Trep6 = MAR2Trep3)

# non-data columns
keep_first  <- colnames(miseq13_vsearch_sum_mod)[1:11]
keep_last   <- tail(colnames(miseq13_vsearch_sum_mod), 2)

# merge all data into a final vsearch data frame for the 10 site study
# Pivot data so it can be bound
miseq13_vsearch_sum_mod_long <- miseq13_vsearch_sum_mod %>%
  pivot_longer(
    cols = -all_of(c(keep_first, keep_last)),  # Specify the range or individual column names
    names_to = "sample",
    values_to = "value"
  ) %>%
  # Ensure that the 'value' column is numeric and any empty strings are replaced with 0
  mutate(value = as.numeric(value)) %>%
  filter(value != 0) %>%  # Optionally filter out rows where value is zero
  select(-matches("^X(\\.\\d+)?$"))  # Remove columns named X, X.1, ..., X.11

miseq15_vsearch_long <- miseq15_vsearch_reruns %>%
  pivot_longer(
   cols = -all_of(c(keep_first, keep_last)),  # Specify the range or individual column names
    names_to = "sample",
    values_to = "value"
  ) %>%
  # Ensure that the 'value' column is numeric and any empty strings are replaced with 0
  mutate(value = as.numeric(value)) %>%
  filter(value != 0) %>%  # Optionally filter out rows where value is zero
  select(-matches("^X(\\.\\d+)?$"))  # Remove columns named X, X.1, ..., X.11
 
miseq16_vsearch_long <- miseq16_vsearch_reruns %>%
  pivot_longer(
  cols = -all_of(c(keep_first, keep_last)),  # Specify the range or individual column names
    names_to = "sample",
    values_to = "value"
  ) %>%
  # Ensure that the 'value' column is numeric and any empty strings are replaced with 0
  mutate(value = as.numeric(value)) %>%
  filter(value != 0) %>%  # Optionally filter out rows where value is zero
  select(-matches("^X(\\.\\d+)?$"))  # Remove columns named X, X.1, ..., X.11

#bind datasets back together
final_10sites_long <- rbind(miseq13_vsearch_sum_mod_long, miseq15_vsearch_long, miseq16_vsearch_long)

# final_10sites_long <- final_10sites_long %>%
#   mutate(sample = str_trim(sample))

final_df  <- final_10sites_long  %>%
  pivot_wider(
    names_from = sample,
    values_from = value,
    values_fill = 0  # optionally fill missing sample-asv combos with 0
  ) 
```



## SECTION 1: Filling in missing/incorrect taxonomic classifications

A file based on analysis of Miseq datasets was constructed ('asv_library_97_aug24.xlsx'). This file captures ASVs (amplicon sequence variants) from the DADA2 step (where the metabarcoding data is reduced into a series of sequence groups with high sequence similarity i.e. ASVs) and this file:
-   fills truncated taxonomy assignments from the vsearch step (sheet 1)
-   groups unclassified sequences into 'OTUs' (sheet 2)
-   updates/corrects any missing identifications or errors in classification (sheet 3)
-   identifies any ambiguous taxonomic assignments that should be considered as complexes (sheet 4)

```{r}
# read in asv library
asv_lib <- read.csv(here::here("data/asv_library_Feb_2026.csv"), 
                    stringsAsFactors = FALSE)

# Replace all NA with blank
asv_lib[is.na(asv_lib)] <- ""
```

# remove identifications below thresholds for species species, genus, family and order
```{r}
# Step 1: Select only taxonomy and match fields from asv_lib
asv_lib_subset <- asv_lib %>%
  select(asv_code, 
         kingdom_new = kingdom, phylum_new = phylum, class_new = class, 
         order_new = order, family_new = family, genus_new = genus, 
         species_new = species, max_p_identity_new = max_p_identity)

# Step 2: Join to final_df and update taxonomy fields ONLY
final_df_updated <- final_df %>%
  left_join(asv_lib_subset, by = "asv_code") %>%
  mutate(
    kingdom        = coalesce(kingdom_new, kingdom),
    phylum         = coalesce(phylum_new, phylum),
    class          = coalesce(class_new, class),
    order          = coalesce(order_new, order),
    family         = coalesce(family_new, family),
    genus          = coalesce(genus_new, genus),
    species        = coalesce(species_new, species),
    max_p_identity = coalesce(max_p_identity_new, max_p_identity)
  ) %>%
  select(-ends_with("_new"))  # remove temporary columns

# Step 6: Recombine with the rest of the final_df that had no update
#final_updated_df <- bind_rows(df1_updated, df3)

# Step 7: Reorder columns if needed
final_df_filled <- final_df_updated %>%
  select(asv_code, kingdom, phylum, class, order, family, genus, species, max_p_identity, everything(), asv_seq, amplicon)
```


```{r}
# Update 'species' column to only include names if max threshold is above 97
final_df_filled <-  final_df_filled %>%
  dplyr::mutate(species = if_else(max_p_identity < 97, "", species))

# Update 'genus' column to only include names if max threshold is above 95
final_df_filled <-  final_df_filled %>%
  dplyr::mutate(genus = if_else(max_p_identity < 95, "", genus))

# Update 'family' column to only include names if max threshold is above 90
final_df_filled <-  final_df_filled %>%
  dplyr::mutate(family = if_else(max_p_identity < 90, "", family))

# Update 'order' column to only include names if max threshold is above 80
final_df_filled <-  final_df_filled %>%
  dplyr::mutate(order = if_else(max_p_identity < 85, "",order))
```


```{r}
# Create results sub-directory inside your project
dir.create(here::here("results/ten_site_data_summaries"), recursive = TRUE, showWarnings = FALSE)

# Add timestamp
timestamp <- format(Sys.time(), "%d_%m_%Y")

# Construct the file path
file_name <- here::here(
  paste0("results/ten_site_data_summaries/ten_sites_vsearch_data_summary_filled_", 
         timestamp, ".csv")
)

# Save the dataframe
write.csv(final_df_filled, file = file_name, row.names = FALSE)
```

____________________________________________________________

# SECTION 2: PCR filtering
This section is designed to format data and filter it for 'mistagging'. Mistagging or 'tag switching' can occur causing ASVs to be assigned to the wrong sample. This occurs at a low frequency and is usually random. PCR replicates included in the study design help detect and remove these errors. We use three PCR replicates and filter out any species/ASV detections that occur in only one PCR replicate (which are most likely from mistagging). We also add a second flitering step for the five most common ASV (those with the highest number of reads/sequences). These are more likely to be involved in mistagging and can appear in multiple PCR replicate we add a reads threshold for removing these detections.

# Change data to long formmat (list) for PCR replicate filtering 
```{r}
# Step 1: Add 'X' prefix to column names if they start with a number
final_df_filled <- final_df_filled %>%
  rename_with(~ str_replace_all(., "^([0-9])", "X\\1"))

# Step 2: Pivot longer to transform selected columns into rows
metab_long <- final_df_filled %>%
  pivot_longer(cols = -(1:13), names_to = "sample", values_to = "value") %>%
  filter(value != 0) %>%  # Remove rows where value is zero
  select(-matches("^X(\\.\\d+)?$"))  # Remove unwanted columns named X, X.1, etc.

# Step 3: Create a 'replicate' column by extracting last character if it ends in 'rep' followed by a number
metab_long <- metab_long %>%
  mutate(replicate = ifelse(grepl("rep[123456]$", sample), str_extract(sample, "rep[123456]$"), ""))

# Step 4: Extract the 'site_per' and 'per' columns based on 'sample'
metab_long <- metab_long %>%
  mutate(site_per = sub("(rep[123456])$", "", sample),  # Remove replicate suffix from sample to get site_per
         per = substr(site_per, 2, 3))  # Take the first two characters of site_per for 'per' column

# Remove the "X" prefix from 'sample' and 'site_per' columns
metab_long <- metab_long %>%
  mutate(
    sample = str_remove(sample, "^X"),
    site_per = str_remove(site_per, "^X")
  )

# Step 5: Create a 'site' column by further trimming 'site_per'
metab_long <- metab_long %>%
  mutate(site = sub("^\\d{2}", "", site_per))  # Remove initial numeric prefix from site_per for site

# Step 6: Reorder and rename columns as needed
metab_long <- metab_long %>%
  select(sample, replicate, site_per, per, site, everything()) %>%  # Adjust column order
  dplyr::rename(reads = value) %>%                                         # Rename 'value' to 'reads'
  mutate(
    replicate = gsub("rep3", "rep1", replicate),
    replicate = gsub("rep4", "rep2", replicate)
  )

# Print the transformed data frame
print(metab_long)
```
```{r}
# Add timestamp
timestamp <- format(Sys.time(), "%d_%m_%Y")

# Construct the file path
file_name <- here::here(
  paste0("results/ten_site_data_summaries/ten_sites_vsearch_data_summary_filled_long_format_", 
         timestamp, ".csv")
)

# Save the dataframe
write.csv(metab_long, file = file_name, row.names = FALSE)
```

We use a final dual filtering strategy:

ASV-level filtering: We first remove potential mistagging by examining the most dominant ASVs. A proportional filter is applied to remove ASVs that are high-frequency globally but appear with a low number of reads in individual samples, as these are most likely to arise from mistagging.

Species-level filtering: We also apply a species-based filter to remove potential mistags originating from dominant species. ASVs with low frequency within these dominant species are filtered out.

Both filtering processes can be adjusted using different thresholds, which are determined based on control samples in the dataset.
```{r}
# PARAMETERS

# Scaled proportional filter
min_thresh <- 0.00
max_thresh <- 0.001

# Dominant species safeguard
species_dom_threshold <- 0.05     # ≥5% of global species reads
min_reads_dom_species <- 11       # minimum reads per ASV for dominant species

# STEP 1 — Remove globally rare ASVs
rows_before_lt5 <- nrow(metab_long)

metab_long_filtered <- metab_long %>%
  filter(reads >= 5)   # keep rows >3 to remove singletons and doubletons

metab_removed_lt5 <- metab_long %>%
  filter(reads < 5)  # rows removed (<3)

rows_after_lt5 <- nrow(metab_long_filtered)

cat("Initial <5 reads filter:\n")
cat("Rows before filter:", rows_before_lt5, "\n")
cat("Rows removed (<=3 reads):", nrow(metab_removed_lt5), "\n")
cat("Rows remaining:", rows_after_lt5, "\n\n")
# STEP 2 — Compute global ASV proportions
asv_global <- metab_long_filtered %>%
  group_by(asv_code, species) %>%
  summarise(total_reads = sum(reads, na.rm = TRUE), .groups = "drop") %>%
  mutate(global_prop = total_reads / sum(total_reads))


# STEP 3 — Compute scaled threshold
asv_global <- asv_global %>%
  mutate(
    scaled_threshold = min_thresh +
      (global_prop / max(global_prop)) *
      (max_thresh - min_thresh)
  )

# STEP 4 — Merge thresholds back to main data
metab_scaled <- metab_long_filtered %>%
  left_join(asv_global %>% select(asv_code, scaled_threshold),
            by = "asv_code")

# STEP 5 — Compute per-sample proportions
metab_scaled <- metab_scaled %>%
  group_by(sample) %>%
  mutate(sample_prop = reads / sum(reads)) %>%
  ungroup()

# STEP 6 — Apply scaled proportional filter
rows_removed_scaled_filter <- metab_scaled %>%
  filter(sample_prop < scaled_threshold)

cat("Scaled ASV filter:\n")
cat("Rows before filter:", nrow(metab_scaled), "\n")
cat("Rows removed:", nrow(rows_removed_scaled_filter), "\n")
cat("Rows remaining:", nrow(metab_scaled) - nrow(rows_removed_scaled_filter), "\n\n")

metab_scaled_filtered <- metab_scaled %>%
  filter(sample_prop >= scaled_threshold)

# STEP 7 — Identify globally dominant species
metab_species_global <- metab_scaled_filtered %>%
  filter(!is.na(species) & species != "") %>%
  group_by(species) %>%
  summarise(total_reads = sum(reads, na.rm = TRUE), .groups = "drop") %>%
  filter(total_reads > 10) %>%
  mutate(global_prop = total_reads / sum(total_reads))

dominant_species <- metab_species_global %>%
  filter(global_prop >= species_dom_threshold) %>%
  pull(species)


# STEP 8 — Apply dominant species + minimum reads filter
rows_removed_dominant_filter <- metab_scaled_filtered %>%
  filter(!is.na(species) & species != "") %>%
  filter(species %in% dominant_species & reads <= min_reads_dom_species)

cat("Dominant species safeguard filter:\n")
cat("Rows removed:", nrow(rows_removed_dominant_filter), "\n")
cat("Rows remaining:", 
    nrow(metab_scaled_filtered) - nrow(rows_removed_dominant_filter), "\n\n")

filtered_metab <- metab_scaled_filtered %>%
  filter(!(species %in% dominant_species & reads <= min_reads_dom_species))


# FINAL SUMMARY
cat("FINAL DATASET:\n")
cat("Total rows:", nrow(filtered_metab), "\n")
cat("Total ASVs:", n_distinct(filtered_metab$asv_code), "\n")
cat("Total samples:", n_distinct(filtered_metab$sample), "\n")
  
```

#Remove control samples to examine elsewhere AppendixA_checking_control
```{r}
pattern_to_remove <- "^(DrosX|MAR|Cont)"

# Controls (rows to remove)
controls <- filtered_metab %>%
  filter(grepl(pattern_to_remove, site))

write.csv(controls,
          here::here("results/final_data_summaries/controls.csv"),
          row.names = FALSE)

# Keep only real samples
filtered_dataframe_samples <- filtered_metab %>%
  filter(!grepl(pattern_to_remove, site) | is.na(site))
```

# save data for modeling in long format (includes PCR replicates)
```{r}
# Generate a timestamp
timestamp <- format(Sys.time(), "%d_%m_%Y")

# Create the file name with the timestamp
file_name1 <- here::here(
  paste0("results/ten_site_data_summaries/ten_sites_final_long_format_", timestamp, ".csv"))

# Save the dataframe with the new file name
write.csv(filtered_dataframe_samples, file = file_name1, row.names = FALSE)
```

