---
title: "macroinvertebrates_morph_comparisions"
author: "Melissa Carew"
date: "11/11/2025"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This notebook is designed to perform analyses on the metropolitan macroinvertebrate component of the dataset. It summarises the reads obtained, the taxa identified, and the macroinvertebrates detected across different sites, including dry weights and sample percentages.

```{r supress warning}
# load libraries
library(tidyverse)
library(vegan)
library(compositions)
library(dplyr)
library(tidyr)
library(ggplot2)
library(RColorBrewer)
library(patchwork)
library(here)

# load required data files

# load site and dry weight data
site_dw <- read.csv(here::here("data/ten_site_dry_weight.csv"))

# Load morphological data files
morph50 <- read.csv(here::here("data/ten_site_50ml_morph_data.csv"))
morph_larg <- read.csv(here::here("data/ten_site_large_frac_data.csv"))

# add taxoncodes to dna data
fambc <- read.csv(here::here("data/family_bugcodes.csv")) # macroinvertebrate taxoncodes

# read in DNA data and align with full data labels (add back in hyphens)
metab_filt <- read.csv(here::here("results/ten_site_data_summaries/ten_sites_final_long_format_24_02_2026.csv")) %>%
  mutate(
    site = case_when(
      site == "FEH41843"  ~ "FEH-4184-3",
      site == "YAR36377"  ~ "YAR-3637-7",
      site == "BNY29042"  ~ "BNY-2904-2",
      site == "UYT1993"   ~ "UYT-199-3",
      site == "WMI2320"   ~ "WMI-232-0",
      site == "FER6475"   ~ "FER-647-5",
      site == "GRD75620"  ~ "GRD-7562-0",
      site == "TOO42700"  ~ "TOO-4270-0",
      site == "DNG318813" ~ "DNG-31881-3",
      site == "DNG79574"  ~ "DNG-7957-4",
      TRUE ~ site  # keep all others unchanged
    )
  )


```

Add in taxoncodes to identify macroinvertebrates
```{r}
metab_filt <- metab_filt %>%
  left_join(fambc %>% select(family, shortcode), 
            by = "family", 
            relationship = "many-to-one") %>%
  mutate(
    # replace NAs from join with empty cells
    shortcode = coalesce(shortcode, ""),
    
    # enforce Eiseniella tetraedra rule
    shortcode = if_else(species == "Eiseniella tetraedra", "LO09", shortcode),
    
        # enforce Macrostomum tuba rule (most like mistaken for Dugusiidae)
    shortcode = if_else(species == "Macrostomum tuba", "IF61", shortcode)
  )

# move data with bug codes into a new df
metab_filt_macro <- metab_filt %>%
  filter(shortcode != "" & !is.na(shortcode))


timestamp <- format(Sys.time(), "%d_%m_%Y")

# Create the file name with the timestamp (NOTE:Change the MiSeq number to the dataset you are working with)
file_name <- here::here(
  paste0("results/ten_site_data_summaries/ten_sites_wide_format_", timestamp, ".csv"))

# truncate code to bioassessment levels for taxa not identified to family
metab_filt_macro_bio <- metab_filt_macro %>%
   mutate(shortcode = case_when(
    str_starts(shortcode, "MO") ~ "MM",
    str_starts(shortcode, "MP") ~ "MM",
    str_starts(shortcode, "LO") ~ "LO",
    TRUE ~ shortcode
  ))

no_bc <- metab_filt %>%
  anti_join(metab_filt_macro, by = "shortcode") #

# Create output directory if needed
dir.create(
  here::here("results/data_analysis/macro_data"),
  recursive = TRUE,
  showWarnings = FALSE
)

write.csv(metab_filt_macro, here::here("results/data_analysis/macro_data/ten_site_metabar_macro.csv"))
```

# summarize macroinvertebrate data detection statistics
How many sequences and ASVs were assigned to species. 
```{r}
# total number of ASVs
num_asv <- length(unique(metab_filt_macro$asv_code))

# short amplicon (BF1/BR1) ASV's 
short_asv <- metab_filt_macro[metab_filt_macro$amplicon == "short", ]
uni_short <- length(unique(short_asv$asv_code))
short_asv_reads <- sum(short_asv$reads)

#right amplicon (mICOIintF/dgHCO2198) ASV's 
right_asv <- metab_filt_macro[metab_filt_macro$amplicon == "right", ]
uni_right <- length(unique(right_asv$asv_code))
right_asv_reads <- sum(right_asv$reads)

# total number of ASVs with species identifications
marcos_97 <- subset(metab_filt_macro, max_p_identity >= 97)
num_asv_sp <- length(unique(marcos_97$asv_code))

# calculate the percentage of ASVs with species ids
percentage_asv_id <- num_asv_sp / num_asv *100

# total number of high quality reads
total_reads <- sum(metab_filt_macro$reads)

# total number of high quality reads with species identifications
sp_reads <- sum(marcos_97$reads)

# calculate the percentage of reads with species identifications
percentage_sp_ids <- sp_reads/ total_reads *100

# how many species/families
sp_count <- length(unique(marcos_97$species))
fam_count <- length(unique(marcos_97$family))

# print stats
print(paste("total_marcoinvertebrate_reads_used in_analysis =", total_reads))
print(paste("Total_no_of__marcoinvertebrate_ASVs =",num_asv))
print(paste("Total_no_of_marcoinvertebrate_ASVs_short amplicon_(B/R5)  =", uni_short))
print(paste("Total_no_of_marcoinvertebrate_reads_short amplicon_(B/R5)  =", short_asv_reads))
print(paste("Total_no_of_marcoinvertebrate_ASVs_right amplicon_(mICOIintF/dgHCO2198)  =", uni_right))
print(paste("Total_no_of_marcoinvertebrate_reads_right amplicon_(mICOIintF/dgHCO2198)  =", right_asv_reads))
print(paste("marcoinvertebrate percentage_asv_id_to_species =", percentage_asv_id))
print(paste("percentage of marcoinvertebrate reads with species identifications", percentage_sp_ids))
print(paste("no. of marcoinvertebrate species = ", sp_count))
print(paste("no. of marcoinvertebrate familes = ", fam_count))
```
Classify all taxa found in run into common name groups to product data for pie charts (Figure 2)
```{r}
# assigning common group names

# --- Assign groups for no_bc ---
no_bc <- no_bc %>%
  mutate(group = case_when(
    # Terrestrial invertebrates
    order %in% c("Hymenoptera", "Araneae", "Sarcoptiformes", "Pseudoscorpiones", "Mesostigmata", "Ixodida", "Trombidiformes") ~ "terrestrial invertebrate",
    class %in% c("Collembola", "Diplopoda", "Chilopoda", "Symphyla") ~ "terrestrial invertebrate",
    phylum %in% c("Mollusca") ~ "terrestrial invertebrate",
    class == "Insecta" & !is.na(species) & species != "" ~ "terrestrial invertebrate",
    family %in% c("Ocnerodrilidae", "Bdellidae", "Lumbricidae") ~ "terrestrial invertebrate",

    # Freshwater microinvertebrates
    class %in% c("Branchiopoda", "Ostracoda", "Hexanauplia", "Gymnolaemata", "Copepoda") ~ "freshwater microinvertebrate",
    phylum %in% c("Tardigrada", "Bryozoa", "Tubulinea", "Gastrotricha") ~ "freshwater microinvertebrate",

    # Fungi
    phylum %in% c("Ascomycota", "Mucoromycota", "Heterokontophyta", "Basidiomycota", "Oomycota", "Zygomycota") ~ "fungi",

    # Fish
    phylum == "Chordata" ~ "fish",

    # Amoeba / Zooplankton / Rotifers
    phylum %in% c("Rotifera", "Amoebozoa", "Evosea", "Discosea", "Cercozoa") ~ "amoeba/rotifers/zooplankton",
    class == "Bigyra" ~ "amoeba/rotifers/zooplankton",
    species == "uncultured zooplankton" ~ "amoeba/rotifers/zooplankton",

    # Algae / Plants
    phylum %in% c("Rhodophyta", "Ochrophyta", "Bacillariophyta", "Chlorophyta", "Streptophyta", "Cryptista", "Gyrista", "Magnoliophyta") ~ "algae/plants",
    class %in% c("Phaeophyceae", "Synurophyceae", "Chrysophyceae") ~ "algae/plants",

    # Unassigned
    kingdom == "Unassigned" ~ "unassigned",
    kingdom == "Eukaryota" & (is.na(phylum) | phylum == "") & (is.na(class) | class == "") ~ "unassigned",

    # Invertebrates (unknown origin)
    phylum %in% c("Platyhelminthes", "Nemertea", "Nematoda", "Cnidaria", "Porifera", "Polychaeta", "invertebrate environmental sample") ~ "invertebrates (unknown origin)",
    class == "Malacostraca" ~ "invertebrates (unknown origin)",
    phylum == "Annelida" & (is.na(species) | species == "") ~ "invertebrates (unknown origin)",
    class %in% c("Insecta", "Trombidiformes") & (is.na(species) | species == "") ~ "invertebrates (unknown origin)",
    species %in% c("Enchytraeida sp. MC-2", "Haplotaxida sp. MC-1") ~ "invertebrates (unknown origin)",
    phylum == "Arthropoda" & (is.na(class) | class == "") ~ "invertebrates (unknown origin)",
    class == "Arachnida" & (is.na(order) | order == "") ~ "invertebrates (unknown origin)",
    order == "Trombidiformes" & (is.na(family) | family == "") ~ "invertebrates (unknown origin)",

    # Bacteria
    kingdom == "Prokaryota" ~ "bacteria",

    # Catch-all fallback
    TRUE ~ NA_character_
  ))

# --- Assign group for metab_filt_macro ---
metab_filt_macro <- metab_filt_macro %>%
  mutate(group = "freshwater macroinvertebrates")

# --- Combine and summarise ---
metab_filt_grps <- rbind(metab_filt_macro, no_bc) %>%
  select(group, asv_code, reads) %>%
  group_by(asv_code, group) %>%
  summarise(reads = sum(reads, na.rm = TRUE), .groups = "drop")

# --- Summarise by group ---
asv_reads_grp_sum <- metab_filt_grps %>%
  group_by(group) %>%
  summarise(
    n_asv = n_distinct(asv_code),
    total_reads = sum(reads, na.rm = TRUE),
    .groups = "drop"
  )

# --- Output ---
dir.create(here::here("results/data_analysis/pie_charts"), recursive = TRUE, showWarnings = FALSE)

write.csv(
  asv_reads_grp_sum,
  here::here("results/data_analysis/pie_charts/asv_reads_grp_sum.csv"),
  row.names = FALSE
)

# View
asv_reads_grp_sum
```
Compare metabarcoding macroinvertebrate families to that in the 50ml sub-sample
```{r}
# summarize number of families in 50mls
morph50_nfam <- morph50 %>%
  group_by(site) %>%
  summarise(nfam_50ml = n_distinct(shortcode))

# summarize abundance of families in 50mls
morph50_abund <- morph50 %>%
  group_by(site) %>%
  summarise(abundance_50ml = sum(count, na.rm = TRUE))

# summarise number of families in large fraction
morph_larg_nfam <- morph_larg %>%
  group_by(site) %>%
  summarise(nfam_larfrac = n_distinct(shortcode))

# summarise abundance in large fraction  
morph_larg_abund <- morph_larg %>% 
  group_by(site) %>%
  summarise(abundance_larfrac = sum(count, na.rm = TRUE))

# --- combine all summaries ---
morph_sum <- morph50_nfam %>%
  full_join(morph50_abund, by = "site") %>%
  full_join(morph_larg_nfam, by = "site") %>%
  full_join(morph_larg_abund, by = "site") %>%
  left_join(
    site_dw %>% select(site, rich_class, abund_class),
    by = "site"
  ) %>%
  select(
    site, rich_class, abund_class,
    nfam_50ml, abundance_50ml,
    nfam_larfrac, abundance_larfrac
  )

# Create output directory if needed
dir.create(
  here::here("results/data_analysis/morph_table"),
  recursive = TRUE,
  showWarnings = FALSE
)

write.csv(morph_sum, here::here("results/data_analysis/morph_table/ten_site_morph_summary.csv"))

print(morph_sum)
```
# Which taxa missing from metabarcoding data - 50mls
```{r}
missing_pairs_morph50 <- morph50 %>%
  filter(!is.na(shortcode), shortcode != "") %>%
  select(site, shortcode, family) %>%
  distinct() %>%   # <-- removes exact duplicates
  anti_join(
    metab_filt_macro_bio %>%
      select(site, shortcode) %>%
      distinct(),
    by = c("site", "shortcode")
  ) %>%
  arrange(site, shortcode)

missing_pairs_morph50
```

# Which taxa missing from metabarcoding data - large fraction
```{r}
missing_pairs_morph_larg <- morph_larg %>%
  filter(!is.na(shortcode), shortcode != "") %>%  # ignore blank or missing shortcodes
  select(site, shortcode, family) %>%
  anti_join(
    metab_filt_macro_bio %>%
      select(site, shortcode) %>%
      distinct(),
    by = c("site", "shortcode")
  ) %>%
  arrange(site, shortcode)

missing_pairs_morph_larg
```
Which species are only in the 10% dry weight?
```{r}
# Step 1: Identify species only in per = 10 % dry weight
species_only_10_per_site <- metab_filt_macro %>%
  group_by(site, species) %>%
  summarize(
    per_values = list(unique(per)),
    n_detections = n(),                          # number of rows
    total_reads = sum(reads, na.rm = TRUE),      # total reads
    .groups = "drop"
  ) %>%
  filter(sapply(per_values, function(x) length(x) == 1 && x == 10)) %>%
  select(site, species, n_detections, total_reads)


# Step 2: Add shortcode information
species_only_10_shortcodes <- species_only_10_per_site %>%
  left_join(
    metab_filt_macro %>% 
      select(species, shortcode) %>% 
      distinct(),
    by = "species"
  )


# Step 3: Check if the shortcode is present in morph_larg at the same site
species_only_10_shortcodes <- species_only_10_shortcodes %>%
  rowwise() %>%
  mutate(
    shortcode_in_morph = shortcode %in% 
      morph_larg$shortcode[morph_larg$site == site]
  ) %>%
  ungroup()


# Keep TRUE matches
species_only_10_shortcodes_true <- species_only_10_shortcodes %>%
  filter(shortcode_in_morph)


# Print summary of reads in these detections
cat("Summary of species only in per=10 and present in morph:\n")
cat("Total rows:", nrow(species_only_10_shortcodes_true), "\n")
cat("Total detections:", sum(species_only_10_shortcodes_true$n_detections), "\n")
cat("Total reads:", sum(species_only_10_shortcodes_true$total_reads), "\n\n")


species_only_10_shortcodes_true

```

Species counts per site and testing for abundance/diversity differences (Figure 3)
```{r}
metab_filt_macro_fam_count <- metab_filt_macro %>%
  group_by(site) %>%
  summarise(shortcode_count = n_distinct(shortcode))

metab_filt_macro_sp_count <- metab_filt_macro %>%
  group_by(site) %>%
  summarise(species = n_distinct(species))

metab_filt_macro_summary <- metab_filt_macro %>%
  group_by(site) %>%
  summarise(
    shortcode_count = n_distinct(shortcode),
    species_count = n_distinct(species)
  )

metab_filt_macro_summary

write.csv(metab_filt_macro_summary, here::here("results/data_analysis/macro_data/metab_filt_macro_summary.csv"))

# Example: if your df is named metab_filt_macro_summary
# with columns: site, species_count, shortcode_count
# rename 'shortcode_count' to 'families' for clarity
metab_filt_macro_summary <- metab_filt_macro_summary %>%
  rename(families = shortcode_count, species = species_count)

# Pivot to long format for ggplot
plot_data <- metab_filt_macro_summary %>%
  pivot_longer(cols = c(species, families),
               names_to = "taxa_type",
               values_to = "count")

# Plot
ggplot(plot_data, aes(x = site, y = count, fill = taxa_type)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_text(aes(label = count),
            position = position_dodge(width = 0.8),
            vjust = -0.3, size = 3.5) +
  scale_fill_manual(values = c("species" = "#4472C4", "families" = "#ED7D31")) +
  labs(x = "Site",
       y = "no. of taxa",
       fill = NULL) +
  theme_minimal(base_size = 13) +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1),
    panel.grid = element_blank(),
    legend.position = "top"
  )

## test for significance

species_family <- plot_data %>%
  select(site, taxa_type, count) %>%
  pivot_wider(
    names_from = taxa_type,
    values_from = count
  )   # gives columns: species, families

rich_abun <- site_dw %>%
  left_join(species_family, by = "site") %>%
  select(site, rich_class, abund_class, species, families)

# t-tests
print(t.test(families ~ rich_class, data = rich_abun))
print(t.test(families ~ abund_class, data = rich_abun))

print(t.test(species ~ rich_class, data = rich_abun))
print(t.test(species ~ abund_class, data = rich_abun))


```
# compare abundance to dry weight
```{r}
abund_dw <- morph_sum %>%
  left_join(site_dw %>% select(site, t_dw), by = "site")

cor.test(abund_dw$abundance_50ml, abund_dw$t_dw, method = "pearson")

# # Your data

# -----------------------------
# 1. Abundance class vs dry weight
# -----------------------------
ggplot(abund_dw, aes(x = abund_class, y = t_dw, fill = abund_class)) +
  geom_boxplot(alpha = 0.6) +
  geom_jitter(width = 0.1) +
  theme_minimal() +
  labs(title = "Dry Weight by Abundance Class")


# -----------------------------
# 2. Abundance vs dry weight
# -----------------------------
ggplot(abund_dw, aes(x = abund_class, y = t_dw)) +
  geom_point(size = 3) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  labs(title = "Abundance vs Dry Weight")


# -----------------------------
# 3. Dry weight vs richness (numeric)
# -----------------------------
ggplot(abund_dw, aes(x = rich_class, y = t_dw, fill = rich_class)) +
  geom_boxplot(alpha = 0.6) +
  geom_jitter(width = 0.1) +
  theme_minimal() +
  labs(title = "Dry Weight by Richness Class")

t.test(t_dw ~ abund_class, data = abund_dw, var.equal = FALSE)

```
`


Box plots of the proportion of species found in each subsample (figure 4)
```{r}
# filter to >=97% identity
metab_filt_macro97 <- metab_filt_macro %>% 
  filter(max_p_identity >= 97)

# Step A: species richness per site × per (what you already had)
site_per_totals <- metab_filt_macro97 %>%
  group_by(site, per) %>%
  summarise(
    total_species = n_distinct(species),
    .groups = "drop"
  )

# Step B: total species per site across all periods (site-level total)
site_totals <- metab_filt_macro97 %>%
  group_by(site) %>%
  summarise(
    site_total_species = n_distinct(species),
    .groups = "drop"
  )

# Step C: join and compute the desired proportion
site_sample_prop <- site_per_totals %>%
  left_join(site_totals, by = "site") %>%
  mutate(
    species_prop = total_species / site_total_species
  )

# Inspect
print(site_sample_prop)

# ---- Example plot: show distribution of species_prop across periods ----
p <- ggplot(site_sample_prop, aes(x = factor(per), y = species_prop)) +
  geom_boxplot(
    fill = "lightblue",
    color = "black",
    width = 0.6,
    #outlier.shape = 19,       # solid black dot
    #outlier.size = 2.5,       # make outlier a bit larger
    #outlier.color = "black"
  ) +
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 18,
    size = 4,
    color = "black"
  ) +
  labs(
    x = "percentage of sample processed",
    y = "proportion of species of total detected",
  ) +
  theme_classic(base_size = 14) +   # Start with a clean theme with borders
  theme(
    panel.background = element_rect(fill = "white", color = "black"),  # white background, black border
    plot.background = element_rect(fill = "white"),                     # background outside panel
    plot.title = element_text(hjust = 0.5))


# save as PNG
ggsave(
  filename = here::here("results/data_analysis/macro_data/species_proportion_plot.png"),
  plot = p,
  width = 8,      # inches
  height = 6,     # inches
  dpi = 300       # high resolution
)

print(p)

anova_prop <- aov(species_prop ~ factor(per), data = site_sample_prop)
summary(anova_prop)

# Summary per period
per_summary <- site_sample_prop %>%
  group_by(per) %>%
  summarise(
    mean_species_prop = mean(species_prop, na.rm = TRUE),
    median_species_prop = median(species_prop, na.rm = TRUE),
    mean_total_species = mean(total_species, na.rm = TRUE),
    median_total_species = median(total_species, na.rm = TRUE),
    .groups = "drop"
  )

per_summary

pno <- ggplot(site_sample_prop, aes(x = factor(per), y = species_prop)) +
  
  geom_boxplot(
    fill = "skyblue",
    color = "black",
    width = 0.6,
    coef = Inf,          #   whiskers to min & max
    outlier.shape = 19,
    outlier.size = 2.5,
    outlier.color = "black"
  ) +
  
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 18,          # diamond
    size = 4,
    color = "black"
  ) +
  
  labs(
    x = "percentage of sample processed",
    y = "proportion of species of total detected"
  ) +
  
  theme_classic(base_size = 14) +
  theme(
    panel.background = element_rect(fill = "white", color = "black"),
    plot.background = element_rect(fill = "white"),
    plot.title = element_text(hjust = 0.5)
  )

# save as PNG
ggsave(
  filename = here::here("results/data_analysis/macro_data/species_proportion_plot_no_outliers.png"),
  plot = pno,
  width = 8,      # inches
  height = 6,     # inches
  dpi = 300       # high resolution
)

print(pno)
```
```{r}
# -------------------------------------------------------------
# Step 0: Filter to ≥97% identity
# -------------------------------------------------------------

metab_filt_macro97 <- metab_filt_macro %>% 
  filter(max_p_identity >= 97)

# -------------------------------------------------------------
# Step 1: Remove species detections with <50 reads
# (sum reads per site × per × species first)
# -------------------------------------------------------------


metab_filt_macro97_50 <- metab_filt_macro97 %>%
  group_by(site, per, species) %>%
  summarise(total_reads = sum(reads), .groups = "drop") %>%
  filter(total_reads >= 5)
# -------------------------------------------------------------
# Step A: species richness per site × per
# -------------------------------------------------------------

site_per_totals <- metab_filt_macro97_50 %>%
  group_by(site, per) %>%
  summarise(
    total_species = n_distinct(species),
    .groups = "drop"
  )

# -------------------------------------------------------------
# Step B: total species per site across all periods
# -------------------------------------------------------------

site_totals <- metab_filt_macro97_50 %>%
  group_by(site) %>%
  summarise(
    site_total_species = n_distinct(species),
    .groups = "drop"
  )

# -------------------------------------------------------------
# Step C: join and compute proportion
# -------------------------------------------------------------

site_sample_prop <- site_per_totals %>%
  left_join(site_totals, by = "site") %>%
  mutate(
    species_prop = total_species / site_total_species
  )

print(site_sample_prop)

pno <- ggplot(site_sample_prop, aes(x = factor(per), y = species_prop)) +
  
  geom_boxplot(
    fill = "skyblue",
    color = "black",
    width = 0.6,
    coef = Inf,          # whiskers to min & max
    outlier.shape = 19,
    outlier.size = 2.5,
    outlier.color = "black"
  ) +
  
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 18,
    size = 4,
    color = "black"
  ) +
  
  labs(
    x = "percentage of sample processed",
    y = "proportion of species of total detected"
  ) +
  
  theme_classic(base_size = 14) +
  theme(
    panel.background = element_rect(fill = "white", color = "black"),
    plot.background = element_rect(fill = "white")
  )

print(pno)
```


```{r}
# -------------------------------------------------------------
# Step 0: Filter to >=97% identity
# -------------------------------------------------------------
metab_filt_macro97 <- metab_filt_macro %>% 
  filter(max_p_identity >= 97)

# -------------------------------------------------------------
# Step 1: Sum reads per site × per × species
# -------------------------------------------------------------
metab_filt_macro97_sum <- metab_filt_macro97 %>%
  group_by(site, per, species) %>%
  summarise(total_reads = sum(reads), .groups = "drop")

# -------------------------------------------------------------
# Step 2: Compute proportion of reads per site × per
# -------------------------------------------------------------
metab_filt_macro97_sum <- metab_filt_macro97_sum %>%
  group_by(site, per) %>%
  mutate(
    reads_prop = total_reads / sum(total_reads) * 100   # % of reads
  ) %>%
  ungroup()

# -------------------------------------------------------------
# Step 3: Filter species below threshold percentage (e.g., 1%)
# -------------------------------------------------------------
threshold_pct <- 1  # adjust as needed, e.g., 1% of total reads
metab_filt_macro97_pct <- metab_filt_macro97_sum %>%
  filter(reads_prop >= threshold_pct)

# -------------------------------------------------------------
# Step 4: Compute species richness per site × per
# -------------------------------------------------------------
site_per_totals <- metab_filt_macro97_pct %>%
  group_by(site, per) %>%
  summarise(
    total_species = n_distinct(species),
    .groups = "drop"
  )

# -------------------------------------------------------------
# Step 5: Compute total species per site across all periods
# -------------------------------------------------------------
site_totals <- metab_filt_macro97_pct %>%
  group_by(site) %>%
  summarise(
    site_total_species = n_distinct(species),
    .groups = "drop"
  )

# -------------------------------------------------------------
# Step 6: Join and compute proportion of species detected
# -------------------------------------------------------------
site_sample_prop <- site_per_totals %>%
  left_join(site_totals, by = "site") %>%
  mutate(
    species_prop = total_species / site_total_species
  )

print(site_sample_prop)

p <- ggplot(site_sample_prop, aes(x = factor(per), y = species_prop)) +
  
  geom_boxplot(
    fill = "skyblue",
    color = "black",
    width = 0.6,
    coef = Inf,          # whiskers to min & max
    outlier.shape = 19,
    outlier.size = 2.5,
    outlier.color = "black"
  ) +
  
  stat_summary(
    fun = mean,
    geom = "point",
    shape = 18,
    size = 4,
    color = "black"
  ) +
  
  labs(
    x = "percentage of sample processed",
    y = "proportion of species of total detected"
  ) +
  
  theme_classic(base_size = 14) +
  theme(
    panel.background = element_rect(fill = "white", color = "black"),
    plot.background = element_rect(fill = "white")
  )

print(p)
```

Community structure between sub-samples (Figure 5)
```{r}
# Step 1: build site_sample × species matrix
comm_matrix <- metab_filt_macro97 %>%
  group_by(site, per, species) %>%
  summarise(abundance = sum(reads), .groups = "drop") %>%
  pivot_wider(names_from = species, values_from = abundance, values_fill = 0)

# Step 2: species matrix only
species_mat <- comm_matrix %>%
  select(-site, -per) %>%
  as.data.frame()

# Step 3: metadata
meta <- comm_matrix %>%
  select(site, per)

# Step 4: CLR transform (pseudocount to avoid log(0))
species_mat_clr <- clr(species_mat + 1) %>%
  as.data.frame()

# Step 5: run ordination (Euclidean distance after CLR = Aitchison distance)
mds <- metaMDS(species_mat_clr, distance = "euclidean", k = 2, trymax = 100)

# Step 6: extract scores
mds_scores <- as.data.frame(scores(mds, display = "sites"))
mds_scores <- cbind(meta, mds_scores)

# Step 7: plot with stress in title
stress_val <- round(mds$stress, 3)


mds_scores$site <- factor(
  mds_scores$site,
  levels = c(
    "HRHA",
    "FEH-4184-3","YAR-3637-7",
    "HRLA",
    "BNY-2904-2","UYT-199-3","WMI-232-0",
    "LRHA",
    "FER-647-5","GRD-7562-0","TOO-4270-0",
    "LRLA",
    "DNG-31881-3","DNG-7957-4"
  ))

 n <- ggplot(mds_scores,
            aes(x = NMDS1,
                y = NMDS2,
                color = site,
                shape = as.factor(per))) +
  geom_point(size = 4, alpha = 0.9) +
  
  scale_color_manual(
    values = site_colors,
    labels = site_labels,
    drop = FALSE
  ) +
  
  labs(
    title = paste0("Stress = ", stress_val),
    x = "NMDS1",
    y = "NMDS2",
    color = "Site",
    shape = "Sample percentage"
  ) +
  
  theme_bw(base_size = 14) +
  theme(
    legend.position = "right",
    panel.border = element_rect(color = "black", fill = NA),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank()
  )
# save as PNG
ggsave(
  filename = here::here("results/data_analysis/macro_data/nmds_precentage_plot.png"),
  plot = n,
  width = 10,      # inches
  height = 6,     # inches
  dpi = 300       # high resolution
)
print(n)
```

Diversity across sample percentage and dry weight (Figure 7)
```{r}
# Make df with species richness and dry weight of sub-samples
site_samp_prop_dw <- metab_filt_macro %>%
  distinct(site, site_per, species, per) %>%      # include 'per'
  group_by(site, site_per, per) %>%
  summarise(
    sp_detected = n(),                            # count species per site_per
    .groups = "drop"
  ) %>%
  group_by(site) %>%
  mutate(
    sp_total = max(sp_detected),                 # total species per site
    sp_prop = sp_detected / sp_total,            # proportion per site_per
    dw_code = str_sub(site_per, 1, 2)           # extract first two digits from site_per
  ) %>%
  ungroup() %>%
  # join DW value for each site_per
  left_join(
    site_dw %>%
      pivot_longer(
        cols = starts_with("dw"),
        names_to = "dw_col",
        values_to = "subsamp_dw"
      ) %>%
      mutate(dw_code = str_remove(dw_col, "dw")) %>%
      select(site, dw_code, subsamp_dw),
    by = c("site", "dw_code")
  ) %>%
  select(-dw_code)  # optional: remove dw_code if you just want subsamp_dw

# Step 1: build site_sample × species matrix
comm_matrix <- metab_filt_macro97 %>%
  group_by(site, per, species) %>%
  summarise(abundance = sum(reads), .groups = "drop") %>%
  pivot_wider(names_from = species, values_from = abundance, values_fill = 0)

# Step 2: species matrix only
species_mat <- comm_matrix %>%
  select(-site, -per) %>%
  as.data.frame()

# Step 3: metadata
meta <- comm_matrix %>%
  select(site, per)

# Step 4: CLR transform (pseudocount to avoid log(0))
species_mat_clr <- clr(species_mat + 1) %>%
  as.data.frame()

# Step 5: run ordination (Euclidean distance after CLR = Aitchison distance)
mds <- metaMDS(species_mat_clr, distance = "euclidean", k = 2, trymax = 100)

# Step 6: extract scores
mds_scores <- as.data.frame(scores(mds, display = "sites"))
mds_scores <- cbind(meta, mds_scores)

# Step 7: plot with stress in title
stress_val <- round(mds$stress, 3)

# -------------------------------------------------------------
# GROUPED SITE LEGEND 
# -------------------------------------------------------------

# Factor and levels for header-style legend
site_samp_prop_dw$site <- factor(
  site_samp_prop_dw$site,
  levels = c(
    "HRHA",
    "FEH-4184-3","YAR-3637-7",
    "HRLA",
    "BNY-2904-2","UYT-199-3","WMI-232-0",
    "LRHA",
    "FER-647-5","GRD-7562-0","TOO-4270-0",
    "LRLA",
    "DNG-31881-3","DNG-7957-4"
  )
)

# Colors (NA for headers)
site_colors <- c(
  NA,
  "#1b9e77","#d95f02",
  NA,
  "#7570b3","#e7298a","#66a61e",
  NA,
  "#e6ab02","#a6761d","#666666",
  NA,
  "#1f78b4","#b2df8a"
)

# Labels with indentation
site_labels <- c(
  "HRHA" = "High richness / High abundance",
  "FEH-4184-3" = "  FEH-4184-3",
  "YAR-3637-7" = "  YAR-3637-7",
  
  "HRLA" = "High richness / Low abundance",
  "BNY-2904-2" = "  BNY-2904-2",
  "UYT-199-3" = "  UYT-199-3",
  "WMI-232-0" = "  WMI-232-0",
  
  "LRHA" = "Low richness / High abundance",
  "FER-647-5" = "  FER-647-5",
  "GRD-7562-0" = "  GRD-7562-0",
  "TOO-4270-0" = "  TOO-4270-0",
  
  "LRLA" = "Low richness / Low abundance",
  "DNG-31881-3" = "  DNG-31881-3",
  "DNG-7957-4" = "  DNG-7957-4"
)

# PANEL A (no legend)
p1 <- ggplot(site_samp_prop_dw, aes(x = per, y = sp_detected, color = site)) +
  geom_line(aes(group = site), linewidth = 0.6, alpha = 0.8) +
  geom_point(size = 2, alpha = 0.9) +
  scale_color_manual(values = site_colors, labels = site_labels, drop = FALSE) +
  scale_x_continuous(labels = function(x) scales::percent(x / 100, accuracy = 1)) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = "sub-sample", y = "species richness") +
  theme_bw(base_size = 12) +
  theme(
    legend.position = "none",
    panel.border = element_rect(color = "black", fill = NA),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank()
  )

# PANEL B (legend only)
p2 <- ggplot(site_samp_prop_dw, aes(x = subsamp_dw, y = sp_detected, color = site)) +
  geom_line(aes(group = site), linewidth = 0.6, alpha = 0.8) +
  geom_jitter(width = 0.15, height = 0.15, size = 2, alpha = 0.9) +
  scale_color_manual(values = site_colors, labels = site_labels, drop = FALSE) +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = "dry weight (g)", y = "species richness", color = "Site") +
  theme_bw(base_size = 12) +
  theme(
    panel.border = element_rect(color = "black", fill = NA),
    panel.grid.minor = element_blank(),
    panel.grid.major = element_blank(),
    legend.position = "right"
  )

# Combine panels (only p2 shows the legend)
combined_plot <- p1 | p2

# Save
ggsave(
  here::here("results/data_analysis/macro_data/species_richness_panels.png"), 
  plot = combined_plot,
  width = 11, height = 4, dpi = 300
)

print(combined_plot)
```




